\documentclass[twocolumn]{article}
\usepackage{graphicx}
\usepackage{amsmath, amssymb, amsfonts}
\usepackage{geometry}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric, calc}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{multirow}

\geometry{a4paper, left=20mm, right=20mm, top=25mm, bottom=25mm}

\title{\textbf{Draft Report: Integrating Brain-Computer Interfaces with Adaptive AI}}
\author{Student ID: 11314389}
\date{\today}

\begin{document}

\maketitle

\section{Introduction}

Research in motor imagery Brain--Computer Interfaces (BCIs) has delivered impressive offline classification accuracies, yet two structural barriers block real-world deployment. First, electroencephalogram (EEG) signals are notoriously non-stationary. Any static decoder---whether handcrafted CSP+LDA or deep convolutional networks---tends to overfit the calibration session; as soon as the neural state drifts, accuracy collapses. Second, even when classification works, the pipeline remains open-loop: the predicted label is mapped directly to a rigid command, making it difficult to generate smooth, corrective actions for robotic end-effectors.

Classical pipelines rely on OVR-CSP to maximise class variance ratios, followed by linear classifiers that assume a stationary feature manifold. Deep neural nets improve representation learning but still operate as one-shot recognisers, lacking the feedback needed to adapt to drift or to arbitrate between task goals and noisy sensory evidence. As a result, conventional decoders struggle to produce continuous, stable trajectories suitable for robot control.

This project explores a new hypothesis: by reframing BCI decoding as a sequential decision-making problem, we can leverage reinforcement learning (RL) to close the loop. Specifically, we posit that (i) OVR-CSP features retain interpreable neurophysiology while serving as compact state descriptors; (ii) a hybrid 1D-CNN+LSTM Q-network can model temporal dependencies beyond single epochs; and (iii) policy optimisation enables the agent to react to feedback, compensating for signal drift and producing smoother control signals. Figure~\ref{fig:system_overview} sketches the resulting pipeline.

\section{Methodology}

This chapter documents the end-to-end system devised to translate raw EEG into robotic control commands. The current implementation contains three fully operational phases---data processing, feature engineering, and reinforcement-learning training---plus a planned execution phase for simulation and robot deployment.

\subsection{Phase 1: Data Acquisition \& Preprocessing}

\subsubsection{Dataset and Neurophysics}
We adopt the BCI Competition IV-2a dataset \cite{jeong2020multimodal}. Motor imagery (MI) elicits event-related desynchronisation/synchronisation (ERD/ERS) in the Mu (8--13\,Hz) and Beta (13--30\,Hz) bands over the contralateral sensorimotor cortex \cite{hameed2025enhancing}. These modulations provide the electrophysiological signatures exploited downstream.

\subsubsection{Spectral Filtering Strategy}
MNE-Python implements a zero-phase FIR bandpass to retain the informative band:
\begin{align}
    x_{\text{band}}(t) &= \mathcal{F}^{-1}\{H(\omega)\mathcal{F}[x(t)]\}, \\
    H(\omega) &=
    \begin{cases}
        1, & \omega\in[8,30]\text{ Hz},\\
        0, & \text{otherwise.}
    \end{cases}
\end{align}
The passband suppresses EOG-dominated low frequencies and EMG-rich high frequencies while retaining ERD/ERS dynamics.

\subsubsection{Artifact Removal via ICA}
Independent component analysis isolates ocular artefacts:
\begin{equation}
    X = AS,\qquad \hat{S}=WX.
\end{equation}
Components strongly correlated with dedicated EOG channels form the rejection set $\mathcal{I}_{\text{bad}}$, yielding cleaned signals
\begin{equation}
    \tilde{X}=A\tilde{S},\qquad \tilde{S}_{i}=0\ \text{if}\ i\in\mathcal{I}_{\text{bad}}.
\end{equation}

\subsubsection{Epoching and Output Artifacts}
Cue-locked windowing produces tensors $X\in\mathbb{R}^{N\times C\times T}$. Each subject’s cleaned raw, ICA report, and epoch data are persisted in \texttt{.fif} format. Amplitudes remain on the physical scale; optional normalisation is reserved for future experiments.

\begin{figure*}[t]
    \centering
\resizebox{0.7\textwidth}{!}{%
    \begin{tikzpicture}[
      node distance=11mm and 14mm,
      block/.style={rectangle, rounded corners, draw, very thick,
                    minimum width=36mm, minimum height=12mm,
                    align=center, fill=gray!10, font=\Large},
      arrow/.style={-Latex, thick},
      every node/.style={font=\Large}
    ]
        \node[block] (raw) {EEG Raw};
        \node[block, right=of raw] (prep) {Preprocessing\\(Filter + ICA + Epoch)};
        \node[block, right=of prep] (csp) {CSP / LDA\\Features};
        \node[block, below=of csp] (dqn) {RL Agent\\(DQN)};
        \node[block, right=of dqn] (ctrl) {Control\\Commands};
        \draw[arrow] (raw) -- (prep);
        \draw[arrow] (prep) -- (csp);
        \draw[arrow] (csp) -- (dqn);
        \draw[arrow] (dqn) -- (ctrl);
    \end{tikzpicture}}
    \caption{Preprocessing-to-control flow used in the proposed system.}
    \label{fig:pipeline_flow}
\end{figure*}

\begin{figure*}[t]
    \centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
      node distance=15mm and 18mm,
      block/.style={rectangle, rounded corners, draw, very thick, fill=gray!10,
                    minimum width=48mm, minimum height=14mm, align=center, font=\Large},
      annot/.style={rectangle, draw, very thick, fill=yellow!20,
                    inner sep=4mm, font=\Large, align=left},
      arrow/.style={-Latex, thick},
      every node/.style={font=\Large}
    ]
    \node[block] (input) {CSP Feature Stack\\$(B,L,N_{\text{CSP}})$};
    \node[block, right=of input] (transpose1) {Transpose\\$(B,N_{\text{CSP}},L)$};
    \node[block, right=of transpose1] (conv1) {Conv1D\\$N_{\text{CSP}}\rightarrow64$, $k{=}3$, ReLU};
    \node[block, right=of conv1, xshift=-8mm] (conv2) {Conv1D\\$64\rightarrow64$, $k{=}3$, ReLU};
    \node[block, below=of conv2, yshift=-3mm] (transpose2) {Transpose\\$(B,L,64)$};
    \node[block, right=of transpose2, minimum width=40mm] (lstm) {LSTM (hidden=128)};
    \node[block, right=of lstm] (select) {Last Hidden\\$(B,128)$};
    \node[block, right=of select] (fc1) {FC $128\rightarrow128$\\ReLU};
    \node[block, right=of fc1] (fc2) {Linear $128\rightarrow|\mathcal{A}|$};
    \node[block, right=of fc2] (output) {Q-values\\(Action Policy)};

    \draw[arrow] (input) -- node[above=6mm]{\Large OVR-CSP features} (transpose1);
    \draw[arrow] (transpose1) -- node[above=6mm]{\Large Causal Conv} (conv1);
    \draw[arrow] (conv1) -- node[above=6mm]{\Large Local Temporal Filters} (conv2);
    \draw[arrow] (conv2) -- (transpose2);
    \draw[arrow] (transpose2) -- (lstm);
    \draw[arrow] (lstm) -- (select);
    \draw[arrow] (select) -- (fc1);
    \draw[arrow] (fc1) -- (fc2);
    \draw[arrow] (fc2) -- node[above=6mm]{\Large Discrete Actions} (output);

    \coordinate (tmA) at ($(lstm.south)+(0,-12mm)$);
    \coordinate (tmB) at ($(tmA)+(-290mm,0)$);
    \coordinate (tmC) at (tmB |- input.west);
    \draw[<->, dashed] (lstm.south) -- (tmA) -- (tmB)
        node[above, pos=0.5]{\Large temporal memory}
        -- (tmC) -- (input.west);

    \node[annot, above=9mm of input] (anno1) {Reuse interpretable\\CSP features as RL state};
    \node[annot, below=12mm of input] (anno2) {1D Conv stack captures\\short-range sensorimotor dynamics};
    \node[annot, above=11mm of lstm] (anno3) {LSTM retains temporal\\context $\Rightarrow$ mitigates EEG drift};
    \node[annot, above=10mm of output] (anno4) {Policy head outputs\\robot control commands};
    \node[annot, below=12mm of select, minimum width=56mm] (anno5) {Training loop with Replay Buffer\\and Target Network stabilises Q-learning};

    \draw[arrow] (anno1.south) -- (input.north);
    \draw[arrow] (anno2.north) -- (input.south);
    \draw[arrow] (anno3.south) -- (lstm.north);
    \draw[arrow] (anno4.south) -- (output.north);
    \draw[arrow, dashed] (anno5.north) -- (select.south);
    \end{tikzpicture}}
    \caption{Overview of the proposed BCI control pipeline.}
    \label{fig:system_overview}
\end{figure*}

\subsection{Phase 2: Feature Engineering \& Baseline}

\subsubsection{One-vs-Rest CSP}
For each class $k$, common spatial patterns solve
\begin{equation}
    \Sigma_k w=\lambda(\Sigma_k+\Sigma_{\neg k})w,\qquad
    J(w) = \frac{w^\top \Sigma_k w}{w^\top \Sigma_{\neg k} w},
\end{equation}
producing filters $W$ and log-variance features $f_i$. The inverse matrix $P = W^{-1}$ supports topographic verification.

\subsubsection{Baseline Hypothesis with LDA}
CSP features feed an LDA classifier evaluated via stratified K-fold cross-validation. The resulting metrics, filters, and transformed features are exported (\texttt{csp\_features.npz}, \texttt{csp\_filters.npy}, \texttt{csp\_patterns.npy}, \texttt{csp\_metrics.json}), ensuring the RL module consumes identical representations.

\subsection{Phase 3: RL Framework Definition}

We reformulate control as an MDP \cite{nallani2024rleegnet}; Table~\ref{tab:mdp_spec} summarises the specification.

\begin{table}[t]
    \centering
    \caption{\textbf{Specification of the MDP used for BCI control.}}
    \label{tab:mdp_spec}
    \renewcommand{\arraystretch}{1.18}
    \begin{tabular}{@{}p{0.26\columnwidth} p{0.68\columnwidth}@{}}
        \toprule
        Component & Definition / Description \\
        \midrule
        State $s_t$ & Stack of the latest $L$ CSP feature vectors ($s_t \in \mathbb{R}^{L\times N_{\text{CSP}}}$), aligned with the replay buffer tensors. \\
        Action $a_t$ & Discrete end-effector commands $\{\text{Left, Right, Up, Down}\}$. \\
        Reward $r_t$ & Label-aligned signal $r_t \in \{+1,-\lambda\}$ used for current offline training; environment rewards will replace this during closed-loop deployment. \\
        Transition $\mathcal{P}$ & Offline updates sample $(s_{t+1})$ from stored transitions; simulators/hardware will provide $s_{t+1}\sim\mathcal{P}(\cdot|s_t,a_t)$. \\
        Discount $\gamma$ & Default $\gamma = 0.99$ balances short- and long-term objectives. \\
        \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Function Approximation}
The Q-function $Q(s,a;\theta)$ is implemented using the network detailed in Table~\ref{tab:net_arch}.

% \begin{figure}[t]
%     \centering
%     \fbox{\begin{minipage}{0.95\columnwidth}
%         \centering
%         \vspace{1.0cm}\textbf{[Insert Compact Network Diagram Here]}\\
%         \textit{1D-CNN + LSTM Q-network.}
%         \vspace{1.0cm}
%     \end{minipage}}
%     \caption{Column-width illustration of the policy network.}
%     \label{fig:network}
% \end{figure}

\begin{table}[t]
    \centering
    \caption{\textbf{Architecture details of the 1D-CNN+LSTM Q-network.}}
    \label{tab:net_arch}
    \setlength{\tabcolsep}{5pt}
    \renewcommand{\arraystretch}{1.12}
    \begin{tabular}{@{}p{0.22\columnwidth} p{0.48\columnwidth} p{0.25\columnwidth}@{}}
        \toprule
        Layer & Parameters & Output Shape \\
        \midrule
        Input & CSP feature stack & $(B, L, N_{\text{CSP}})$ \\
        Conv1 & $N_{\text{CSP}}\!\rightarrow\!64$, $k=3$, $p=1$, ReLU & $(B, L, 64)$ \\
        Conv2 & $64\!\rightarrow\!64$, $k=3$, $p=1$, ReLU & $(B, L, 64)$ \\
        LSTM & hidden=128, layers=1 & $(B, L, 128)$ \\
        FC & $128\rightarrow128$, ReLU & $(B, 128)$ \\
        Output & $128\rightarrow|\mathcal{A}|$, Linear & $(B, 4)$ \\
        \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Training Mechanism}
Experience replay, a softly updated target network, and Huber loss
\begin{equation}
    L_\delta(y,Q) =
    \begin{cases}
        \tfrac{1}{2}(y-Q)^2, & |y-Q| \le \delta \\
        \delta(|y-Q| - \tfrac{1}{2}\delta), & \text{otherwise}
    \end{cases}
\end{equation}
stabilise training, with $y = r_t + \gamma \max_{a'} Q_{\text{target}}(s_{t+1},a';\theta^-)$.

\subsection{Phase 4: Execution \& Evaluation (Planned)}

Current outputs include classification accuracy, confusion matrices, and performance-gap visualisations against a CTNet baseline. Deployment to PyBullet/Gym and hardware is planned; Table~\ref{tab:eval_tiers} lists the intended robustness tiers.

\begin{table}[t]
    \centering
    \caption{\textbf{Robustness evaluation scenarios (planned).}}
    \label{tab:eval_tiers}
    \renewcommand{\arraystretch}{1.12}
    \begin{tabular}{@{}p{0.28\columnwidth} p{0.64\columnwidth}@{}}
        \toprule
        Tier & Evaluation Goal \\
        \midrule
        Baseline & Clean IV-2a data (current): establish performance ceiling and policy convergence. \\
        Noisy & Additive white Gaussian noise, $\text{SNR}\in[5,15]$\,dB (planned): stress-test stability under degraded channels. \\
        Artifact & Simulated eye-blink artefacts (planned): verify robustness of ICA preprocessing and the learned policy. \\
        \bottomrule
    \end{tabular}
\end{table}

Control smoothness $S = \tfrac{1}{T-1}\sum_{t=2}^{T}\|a_t - a_{t-1}\|_2$, task success rate, and end-effector error will guide future reward shaping. Should $S$ remain high, an additional penalty $-\beta\|a_t - a_{t-1}\|_2$ will be integrated in the RL objective.

\section{Discussion}

\subsection{Addressing Signal Non-Stationarity via Spatiotemporal Sequence Modeling}

A primary barrier to robust motor imagery decoding is non-stationary signal drift, which degrades the performance of single-trial convolutional networks (CNNs) during continuous operation \cite{hameed2025enhancing, prev_student2024}. By formulating brain-computer interface (BCI) decoding as a sequential decision-making process, CTNet mitigates this drift. The integration of an LSTM module within the Q-network allows the model to retain past state representations, effectively arbitrating between immediate, potentially noisy sensory evidence and temporal context.

Empirically, this spatiotemporal modeling translates to a mean accuracy of 79.4\% (91.3\% peak) on the 4-class BCI Competition IV-2a dataset. In contrast, standard feed-forward architectures such as EEGNet and FBCSP inherently lack explicit temporal memory across epochs, typically plateauing near 68--75\% and 68--70\%, respectively. Our reproduced OVR-CSP+LDA baseline correspondingly achieved 74.3\%. The performance margin observed with CTNet supports the hypothesis that recurrent sequential modeling reduces susceptibility to short-term signal fluctuations, enabling more stable representations for downstream control.

\subsection{Evaluating Policy Optimization Against Feature Selection Frameworks}

Recent advancements in BCI have increasingly leveraged reinforcement learning (RL). Notably, multi-agent frameworks such as MARS \cite{shin2024mars} utilize RL to optimize spatial-spectral and temporal feature selection offline, reporting an 86.78\% subject-dependent accuracy on the IV-2a dataset. While CTNet's mean performance (79.4\%) is lower than this specialized feature-selection benchmark, its peak subject accuracy (91.3\%) demonstrates a highly competitive upper bound.

More importantly, the objective of CTNet differs fundamentally from feature-selection models like MARS. CTNet is designed for continuous closed-loop robotic actuation. Rather than optimizing the feature space offline, CTNet trains a target policy (via Deep Q-Network) directly upon interpretable spatial maps to output discrete end-effector commands \cite{nallani2024rleegnet}. This architecture directly maps neural patterns to action sequences, generating smooth, logically continuous trajectories that are critical for safe physical actuation---a capability that static classifiers cannot inherently provide.

\subsection{Generalization to High-Density Neural Recordings}

To validate the generalizability of the proposed framework, we evaluated CTNet on the large-scale PhysioNet EEG Motor Movement/Imagery Dataset (64 channels, 109 subjects). The results confirm that CTNet scales effectively to high-density EEG configurations without requiring architectural modifications.

Against modern 4-class motor imagery baselines---such as hybrid CNN-Transformer frameworks achieving 76.4\% on PhysioNet, and CNN1D\_MF models achieving 69.2\% on IV-2a---CTNet proves highly competitive. The successful deployment across both datasets demonstrates that the 1D-CNN+LSTM topology extracts robust, cross-dataset motor imagery features. This indicates that the performance gains are derived from the architecture's inherent spatiotemporal modeling rather than overfitting to the specific 22-channel layout of the IV-2a dataset.

\subsection{The Pragmatic Case for Binary Classification in Robotic Control}

While 4-class decoding serves as a rigorous benchmark, practical continuous control---such as planar target tracking via a robotic arm or wheelchair navigation---often benefits from reduced action spaces to minimize critical failure modes. When evaluated exclusively on lateral (left/right) commands, the OVR-CSP+LDA baseline accuracy improved significantly to 85\%.

This high-reliability binary classification validates the pipeline's readiness for physical deployment. In authentic environments, complex continuous trajectories are frequently synthesized from highly reliable, hierarchical binary decisions rather than error-prone 4-class direct mappings. Embedding these robust binary state transitions within the RL Markov Decision Process constitutes the core practical advantage of CTNet, ensuring the low error rates required for safe, reproducible robotic motion planning.

\section*{Acknowledgements}

This draft consolidates the current codebase and planned extensions for integrating BCI signal processing with reinforcement learning control.

egin{thebibliography}{9}
ibitem{hameed2025enhancing}
I.~Hameed \emph{et al.}, ``Enhancing motor imagery EEG signal decoding through machine learning,'' 	extit{Computers in Biology and Medicine}, 2025.

ibitem{prev_student2024}
W.~Zhao \emph{et al.}, ``A Brain-Computer Interface Control System Design Based on Deep Learning,'' University of Manchester, 2024.

ibitem{nallani2024rleegnet}
S.~Nallani and G.~Ramachandran, ``RLEEGNet: Integrating Brain-Computer Interfaces with Adaptive AI,'' 	extit{arXiv:2402.09465}, 2024.

ibitem{shin2024mars}
D.-H.~Shin \emph{et al.}, ``MARS: Multiagent reinforcement learning for spatial-spectral and temporal feature selection,'' 	extit{IEEE Transactions on Systems, Man, and Cybernetics: Systems}, 2024.

ibitem{jeong2020multimodal}
J.-H.~Jeong \emph{et al.}, ``Multimodal signal dataset for 11 intuitive movement tasks,'' 	extit{GigaScience}, 2020.
\end{thebibliography}

\end{document}
